---
title: "Where a distribution puts its probability"
date: '2026-09-18'
slug: where-a-distribution-puts-its-probability
categories: []
tags:
- probaverse
- R
- CRAN
- probability distributions
- distribution objects
- distribution manipulation
image:
  filename: probaverse-1716x1997.png
  focal_point: Smart
  alt_text: probaverse package logo
  preview_only: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

Three releases went to CRAN this week, and they are really one idea, plus a new verb: the graft.

- **distionary 0.2.0** gives every distribution a *support*.
- **distplyr 0.3.0** rebuilds every manipulation verb on top of supports.
- **probaverse 0.1.1** pins the suite so that installing it gets you the matching set.

## A distribution knew *how*, but not *where*

A distionary distribution has always known *how* to be evaluated. Hand it a number and it hands back a CDF value, a density, a quantile. What it did not know was *where* it puts discrete probability.

For the built-in families, that hardly mattered, because each one arrives with its own formulas. The trouble starts when a distribution is not one of the built-ins --- when you have defined it yourself, or transformed one into something new. Then anything the package has to *derive*, like a quantile it must solve for, needs some assumption about where discrete probability sits. The assumption was: continuous, spread smoothly over an interval.

So a distribution now carries a support: the set where its probability lives, with atoms tracked separately from continuous regions.

```{r}
library(probaverse)
support(dst_pois(3))
```

Those atoms are not a vector that I quietly cut off at some large number. It is the entire infinite set, described by the rule that generates it, which is what the [discretes package](/post/2026-03-31-introducing-discretes-0-1-0/) was built to make possible.

The payoff arrives as soon as mass and density share a distribution. Mix a point mass at zero with an exponential --- 40% dry days, rain otherwise --- and the quantiles land on zero exactly when they should, rather than "roughly zero":

```{r}
rain <- mix(dst_degenerate(0), dst_exp(rate = 0.2), weights = c(0.4, 0.6))
support(rain)
eval_quantile(rain, at = c(0.3, 0.4, 0.5))
```

## Grafting a tail onto history

The more interesting case is the new graft verbs, where supports really earn themselves.

Historical Simulation --- using the observed record directly as the distribution --- is about the most honest thing you can do with data, right up until you ask it about something rarer than anything you have seen. Here are the hundred annual flows of the Nile:

```{r}
x <- as.numeric(Nile)
flows <- dst_empirical(x)
vtype(flows)
eval_quantile(flows, at = 0.999)
```

Discrete, with an atom on every observation --- and its answer to a one-in-a-thousand question is just the largest flood on record. It has nowhere else to put one.

`graft_right()` replaces the top of a distribution with a model of the top. Fit a generalised Pareto to the exceedances over 1150 with famish, then graft it on. The new `knot` argument decides which side the flow sitting exactly on 1150 goes to, and by default it stays with the body:

```{r}
u <- 1150
flood <- graft_right(flows, of = u, tail_excess = fit_dst_gp(x[x > u] - u))
flood
```

The result is mixed, and the support says precisely what happened to it: the observations below the knot survive as atoms --- including 1150 itself, listed last --- and above it there is density instead. The fitted tail came out with a negative shape, so it is bounded --- and the support is where that upper limit shows up, rather than something you work out by hand afterwards.

```{r}
support(flood)
eval_quantile(flood, at = c(0.99, 0.999))
```

None of this is a special case that distplyr hard-codes. The verb asks the empirical distribution where its probability is, maps it, and hands back something that can answer the same question about itself.

## One install, one working set

The catch with a suite is that a good idea in one package is a breaking change in the others: distplyr 0.3.0 genuinely requires distionary 0.2.0. So probaverse 0.1.1 does something dull and worthwhile, and records minimum versions for the packages it loads.

```r
install.packages("probaverse")
```

That now brings a mixed-up library back up to a set that works together, rather than leaving you to find out which pair disagrees.

## What this opens up

When distionary [went to CRAN](/post/2025-12-01-distionary-launch/), I wrote that explicit support for discrete and mixed distributions was coming soon. This is that, and it took a detour through a package about infinite sets to get here.

Real systems are lumpy. A flow that is sometimes zero, an empirical distribution that is nothing but discrete points --- until now the probaverse could represent these but could not reliably compute with them. Now the lump is a stated fact about the distribution that every routine can read, including routines I have not written yet.

Next on the roadmap: multivariate distributions, and distribution families rather than single distributions.
