---
title: "Where a distribution puts its probability"
date: '2026-09-18'
slug: where-a-distribution-puts-its-probability
categories: []
tags:
- probaverse
- R
- CRAN
- probability distributions
- distribution objects
- distribution manipulation
image:
  filename: probaverse-1716x1997.png
  focal_point: Smart
  alt_text: probaverse package logo
  preview_only: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

Three releases went to CRAN this week, and they are really one idea.

- **distionary 0.2.0** gives every distribution a *support*.
- **distplyr 0.3.0** rebuilds every manipulation verb on top of supports.
- **probaverse 0.1.1** pins the suite so that installing it gets you the matching set.

Here is the idea, and why I think it is the most useful thing to land in the probaverse so far.

## A distribution knew *how*, but not *where*

A distionary distribution has always known *how* to be evaluated. Hand it a number and it hands back a CDF value, a density, a quantile. What it did not know was *where* it puts its probability.

For the built-in families that hardly mattered, because each one arrives with its own formulas. The trouble starts when a distribution is not one of the built-ins --- when you have defined it yourself, or transformed one into something new. Then anything the package has to *derive*, like a quantile it must solve for or a moment it must integrate, needs some assumption about where the probability sits. The assumption was: continuous, spread smoothly over an interval.

That is a fine assumption right up until it isn't. Think about daily rainfall. Most days are dry, so the record is mostly zeros. This is not a continuous distribution that happens to have a small mean; it is a lump of probability sitting exactly on zero, with a continuous curve off to the right of it. The same shape of problem shows up in an insurance portfolio where most policies never claim, or a river reach that stops flowing in late summer.

That lump is the whole difficulty. Ask a continuity assumption for the median and it goes hunting for the place where the CDF crosses 0.5, sliding around a jump where no such crossing exists.

## Supports

So a distribution now carries a support: the set where its probability lives, with atoms tracked separately from continuous regions.

```{r}
library(probaverse)
support(dst_pois(3))
```

Those atoms are not a vector that I quietly cut off at some large number. It is the entire infinite set, described by the rule that generates it, which is what the [discretes package](/post/2026-03-31-introducing-discretes-0-1-0/) was built to make possible.

Now for the rainfall problem. Say 40% of days are dry and the wet days follow an exponential distribution. The new `.support` argument is how you say that out loud:

```{r}
p <- 0.4
rain <- distribution(
  cdf = function(x) ifelse(x < 0, 0, p + (1 - p) * pexp(x, rate = 0.2)),
  density = function(x) ifelse(x <= 0, 0, (1 - p) * dexp(x, rate = 0.2)),
  pmf = function(x) ifelse(x == 0, p, 0),
  .support = mixed(discrete = 0, continuous = c(0, Inf)),
  .name = "Zero-inflated Exponential"
)
support(rain)
```

Worth being careful about what went in there. A mixed distribution has no density of its own --- you cannot differentiate a jump --- so the `density` argument carries the *continuous part* alone, the `pmf` carries the mass sitting on zero, and the support is what tells the package how the two fit together. That division is the whole trick: neither piece is the distribution, and previously there was nowhere to write down how they joined.

## So what?

The payoff is that the derived quantities stop being approximate, or stop erroring outright.

Quantiles land *on* the atom rather than near it, along a path that used to raise an error outright for a discrete or mixed distribution. Any probability up to 0.4 is answered with exactly zero, because that is the honest answer --- the driest 40% of days had no rain at all:

```{r}
eval_quantile(rain, at = c(0.2, 0.4, 0.5, 0.9))
```

Moments work too, by summing over the atoms and integrating over the regions --- a calculation the package simply did not do for anything but a continuous distribution before:

```{r}
mean(rain)
variance(rain)
```

And because distplyr 0.3.0 reads and rewrites supports rather than guessing at them, a manipulated distribution still knows where it lives. Two of the new verbs, `trim_left()` and `graft_right()`, take a `knot` argument that decides what happens to mass sitting exactly on the cut point --- which is precisely the question the old machinery could not even ask. Trim away the dry days and what remains is no longer mixed at all:

```{r}
wet <- trim_left(rain, of = 0, knot = "discard")
support(wet)
mean(wet)
```

A mean of 5 mm on wet days, from a distribution averaging 3 mm over all days. Same model, and now both numbers are available from it without my having to take the thing apart by hand.

## One install, one working set

The catch with a suite of packages is that a good idea in one of them is a breaking change in the others. distplyr 0.3.0 genuinely requires distionary 0.2.0; there is no meaningful way to run the new verbs on the old distribution object. So probaverse 0.1.1 does something dull and worthwhile: it records minimum versions for the packages it loads.

```r
install.packages("probaverse")
```

That now brings a library with a mixed-up suite in it back up to a set that works together, rather than leaving you to find out which pair disagrees.

## What this opens up

When distionary [went to CRAN](/post/2025-12-01-distionary-launch/), I wrote that explicit support for discrete and mixed distributions was coming soon. This is that, and it took a detour through a package about infinite sets to get here.

It matters beyond tidier quantiles. Real systems are lumpy. A flow that is sometimes zero, a loss that is usually nothing, a measurement that piles up at a detection limit --- these are mixed distributions, and until now the probaverse could represent them but could not reliably compute with them. Now the lump is a stated fact about the distribution that every routine can read, including routines I have not written yet.

Next on the roadmap: multivariate distributions, and distribution families rather than single distributions. As always, how fast that arrives depends on funding.
